State and Parameter Estimation for Spatio-Temporal Neural Fields

### 2010-05-02 ltx/Journal of Neuroengineering/StateAndParameterEstimationNeuralFields_JoNE.tex
parameter distribution graphs added to the paper
restored Dean's intro structure from an old commit.

\subsection{The Importance of Cortical Structure}

Neural field models relate mean firing rates of pre-synaptic neural populations to post-synaptic mean membrane potentials. Each neural population represents a functional cortical processing unit, referred to as a column. The columnar organization of the cortex is continuous, where pyramidal cells are members of many columns. In general, cortical structure can be modeled in a physiologically plausible manner as being locally homogeneous (in short range intracortical connectivity) and heterogeneous (in long range cortico-cortical and corticothalamic connectivity) \cite{Jirsa2009} \cite{Qubbaj2007}. Locally, each column is thought be connected via symmetric short range local excitation, with surround inhibition \cite{Braitenberg1998}. For example, this structural organisation is most studied in the visual system, where the surrounding inhibition effectively tunes a cortical column to a particular receptive visual field \cite{Sullivan2006}.

Recent studies using neural field models have demonstrated the theoretical implications of specific connectivity structure of the cortical columns in neural field models, where the connectivity kernel governs the bifurcation points of the system \cite{Hutt2005} and types of oscillations that can be generated \cite{Schmidt2009}. This implies that if we could estimate the connectivity structure for an individual, then we could capture patient specific neurodynamics that lead to various oscillatory states.

Estimating functional cortical connectivity via EEG promises to shed light on the hierarchy of brain regions that are involved in specific tasks, motivating the use of techniques like DCM. Other comparable methods are based on auto-regressive (AR) modelling of the EEG, MEG, or fMRI times series \todo{[refs!]}. Studies have concentrated on finding functional connectivity patterns using information contained in the AR coefficients using techniques like Granger causality \cite{Hesse2003}, the direct transfer function \cite{Kaminski1991}, and partial directed coherence \cite{Sameshima1999}. Again, like the DCM approach, these methods estimate long range connectivity patterns that does not provide a clear relationship between the continuum field models and data.

%Analysis of the connectivity estimates provided from these approaches has involved graph theoretical techniques (small worldedness).
Until now, estimation of local intracortical connectivity structure has not been attempted. Recently, it has been shown that it is possible to estimate local coupling of spatiotemporal systems using techniques from control systems theory and machine learning \cite{Dewar2009}. The key development in this work was to represent the spatiotemporal system as a standard state-space model with an order independent of the number of sensors (ECoG recording electrodes in this case). In addition, the appropriate model selection tools have been developed \cite{Scerri2009} allowing for the application of this technique to neural fields. Modelling the neural dynamics within this framework has the distinct advantage over multivariate AR models, that the number of parameters to define the spatial connectivity is considerably smaller than the number of AR coefficients typically required to achieve the required model complexity. In this paper, we demonstrate for the first time how intracortical connectivity can be inferred from ECoG data, based on a variant of the Wilson and Cowan neural field model \cite{Wilson1973}. This work provides a fundamental link between the theoretical advances in neural field modelling and patient specific data.

%Long range connectivity can be estimated using DTI \cite{Knock2009}, however this short range connectivity can not..
\section{Neural Field Model}
\label{NeuralModelSection}

In this section, we describe a variant of the WC neural field model \cite{Wilson1973} that was used to create our intracortical connectivity estimator. This model is descriptive of a range of neurodynamics of the cortex such as evoked potentials, visual hallucinations, and epileptic behaviour. 
\begin{todo}
	{Really need to add some citations here.} 
\end{todo}
It is also capable of generating complex patterns of activity such as Turing patterns, spirals, and traveling oscillations. 
\begin{todo}
	{Same here.} 
\end{todo}

@@

The state estimation technique utilises the Unscented Rauch Tung Striebel Smoother (URTSS)~\cite{Sarkka2010}. The URTSS uses the Unscented Kalman Filter (UKF)~\cite{Julier1997,Merwe2003} in the forward iteration to estimate posterior states, $\hat{\mathbf x}_t^{f}$, followed by a backward path to compute the smoothed state estimates, $\hat{\mathbf x}_t^{b}$. An approximation of the state density is made by propagating the so-called sigma points through the state-space model. The sigma points describe the mean and covariance of the state estimate and are calculated using the unscented transform. The sigma vectors $\mathcal X_i$ are calculated by 
\begin{eqnarray}\label{eq:sigmapoints1}
	\mathcal X_{0}&=&\bar x \\
	\mathcal X_{i}&=&\bar x+(\sqrt{( L + \lambda)\mathbf P_x})_i \quad i=1, \dots, L \\
\label{eq:sigmapoints2}
	\mathcal X_{i}&=&\bar x-(\sqrt{( L + \lambda)\mathbf P_x})_{i- L} \quad i= L+1, \dots, 2 L \nonumber \\
\end{eqnarray}
where $(\sqrt{( L + \lambda)\mathbf P_x})_i$ is the ith column of the weighted matrix squre root of  $\mathbf P_x$ and  $L$ is the dimension of the state space. The total number of sigma points is $2L+1$. The scaling parameter $\lambda$ is defined as 
\begin{equation}\label{eq:sigmapoints3}
	\lambda=\alpha^2( L+\kappa)- L 
\end{equation}
where $\alpha$, $\beta$ and $\kappa$ are constants (see ~\cite{Haykin2001} for more details). It should be noted that $\bar x$ represents either $\hat{\mathbf x}_t^{f}$ or $\hat{\mathbf x}_t^{b}$ and $\mathbf{P}_x$ represents the corresponding covariance matrix from the filtering or smoothing respectively. The sigma vectors are propagated through the system equations and weighted to form the posterior mean and covariance. The weights are calculated by 
\begin{equation}
	\mathbf W_0^{(m)}=\frac{\lambda}{ L+\lambda} 
\end{equation}
\begin{equation}
	\mathbf W_0^{(c)}=\frac{\lambda}{ L+\lambda}+(1-\alpha^2+\beta) 
\end{equation}
\begin{equation}
	\mathbf W_i^{(m)}=\mathbf W_i^{(c)}=\frac{1}{2( L+\lambda)} \quad i=1, \dots, 2L. 
\end{equation}

>>>>

The paper proceeds by first deriving the continuum neural field equations in Section~\ref{NeuralModelSection}. Next, a reduced finite dimensional neural field model is derived in Section~\ref{ReducedModelSection}. The model is reduced by approximating the neural field using a set of continuous basis functions, weighted by a finite dimensional state vector. Section~\ref{SpectralAnalysisSection} then establishes conditions using spatial frequency analysis for both sensor and basis function spacing and geometry, such that the dominate dynamics of the neural field can be represented by the reduced model. Following this, the state and parameter estimation procedure is described in Section~\ref{StateAndParameterEstimationSection}. The results for the spatial frequency analysis and parameter estimation are then presented in Section~\ref{ResultsSection}. The implications and limitations of this framework are discussed in Section~\ref{DiscussionSection} along with planned future developments. Finally, we provide concluding remarks in Section~\ref{ConclusionSection}.

\section{Neural Field Model}\label{NeuralModelSection} In this section, we describe a variant of the WC neural field model \cite{Wilson1973} that was used to create our intracortical connectivity estimator. Neural field models relate mean firing rates of pre-synaptic neural populations to mean post-synaptic membrane potentials. They are popular due to being parsimonious yet having a strong link with the underlying physiology. Each neural population represents a functional cortical processing unit, such as a column. The columnar organisation of the cortex is continuous, where pyramidal cells are members of many columns. In general, cortical structure can be modelled in a physiologically plausible manner as being locally homogeneous (in short range intracortical connectivity) and heterogeneous (in long range cortico-cortical and corticothalamic connectivity)~\cite{Jirsa2009}~\cite{Qubbaj2007}. Locally, each column is thought be connected via symmetric short range local excitation, with surround inhibition \cite{Braitenberg1998}. For example, this structural organisation is most studied in the visual system, where the surrounding inhibition effectively tunes a cortical column to a particular receptive visual field~\cite{Sullivan2006}. Neural field models are descriptive of a range of neurodynamics of the cortex such as evoked potentials, visual hallucinations, and epileptic behaviour. 
\begin{todo}
	{Really need to add some citations here.} 
\end{todo}
It is also capable of generating complex patterns of activity such as Turing patterns, spirals, and traveling oscillations. 
\begin{todo}
	{Same here.} 
\end{todo}

@@

The state estimation technique utilises the Unscented Rauch Tung Striebel Smoother (URTSS)~\cite{Sarkka2010}. The URTSS uses the Unscented Kalman Filter (UKF)~\cite{Julier1997,Merwe2003} in the forward iteration to estimate posterior states, $\hat{\mathbf x}_t^{f}$, followed by a backward path to compute the smoothed state estimates, $\hat{\mathbf x}_t^{b}$. An approximation of the state density is made by propagating the so-called sigma points through the state-space model. The sigma points describe the mean and covariance of the state estimate and are calculated using the unscented transform. The sigma vectors $\mathcal X_i$ are calculated by 
\begin{equation}\label{eq:sigmapoints1}
	\mathcal X_{0}=\bar x 
\end{equation}
\begin{equation}
	\mathcal X_{i}=\bar x+(\sqrt{( L + \lambda)\mathbf P_x})_i \quad i=1, \dots, L 
\end{equation}
\begin{equation}\label{eq:sigmapoints2}
	\mathcal X_{i}=\bar x-(\sqrt{( L + \lambda)\mathbf P_x})_{i- L} \quad i= L+1, \dots, 2 L 
\end{equation}
where $(\sqrt{( L + \lambda)\mathbf P_x})_i$ is the ith column of the weighted matrix squre root of  $\mathbf P_x$ and  $L$ is the dimension of the state space. The total number of sigma points is $2L+1$. The scaling parameter $\lambda$ is defined as 
\begin{equation}\label{eq:sigmapoints3}
	\lambda=\alpha^2( L+\kappa)- L 
\end{equation}
where $\alpha$, $\beta$ and $\kappa$ are constants (see ~\cite{Haykin2001} for more details). It should be noted that $\bar x$ represents either $\hat{\mathbf x}_t^{f}$ or $\hat{\mathbf x}_t^{b}$ and $\mathbf{P}_x$ represents the corresponding covariance matrix from the filtering or smoothing respectively. The sigma vectors are propagated through the system equations and weighted to form the posterior mean and covariance. The weights are calculated by 
\begin{equation}
	\mathbf W_0^{(m)}=\frac{\lambda}{ L+\lambda} 
\end{equation}
\begin{equation}
	\mathbf W_0^{(c)}=\frac{\lambda}{ L+\lambda}+(1-\alpha^2+\beta) 
\end{equation}
\begin{equation}
	\mathbf W_i^{(m)}=\mathbf W_i^{(c)}=\frac{1}{2( L+\lambda)} \quad i=1, \dots, 2L. 
\end{equation}
